k-plus Anticlustering: An Improved k-means Criterion for Maximizing Between-Group Similarity

### 2023-05-05 paper.Rmd
another round of editing

Maximizing $\mathit{SSE_{kplus}}$ will lead to similar means and variances between $K$ groups, while the weights $w_1$ and $w_2$ can be chosen to adjust the relative importance of equalizing means and variances, respectively.

@@

Research on multiobjective optimization indicates that solutions oftentimes fail to satisfactorily address several criteria if the underlying objectives are in conflict (Ferligoj & Batagelj, 1992). As $\mathit{SSE_{kplus}}$ is a novel criterion, it is so far unknown whether maximizing it can yield a partitioning where both constituting criteria are similar between groups, i.e., whether these objectives are aligned or in conflict. A motivating example illustrates that aiming to simultaneously optimize similarity of means and variances may be feasible. I created 14 data points following a univariate normal distribution ($M = 15$, $SD = 3$) and then generated all `r n_partitions(14, 2)` possible ways to partition the 14 data points into $K = 2$ equal-sized groups. Figure 1 shows the scatterplot of the criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ across partitions. The partition that is represented by the squared symbol has the overall best highest $\mathit{SSE}$, but is outperformed by many other partitions with regard to $\mathit{SSE_{Var}}$. Similarly, the partition that maximizes $\mathit{SSE_{Var}}$ (represented by the triangular symbol) is not well-suited to optimize $\mathit{SSE}$ at the same time. Thus, when only considering one of the objectives, the mathematically optimal partitioning fails to establish groups that are similar both with regard to their means and variances. The diamond symbol represents the partition that maximizes the unweighted sum of $\mathit{SSE_{Var}}$ and $\mathit{SSE}$ (i.e., $w_1$ = $w_2$ = 1 in the computation of $\mathit{SSE_{kplus}}$), which yields `r df$kmeans[which.max(df$kplus)] / max(df$kmeans) * 100`% of the global maximum $\mathit{SSE}$, and `r df$kvar[which.max(df$kplus)] / max(df$kvar) * 100`% of the global maximum value of $\mathit{SSE_{Var}}$. Thus, even using a very simple unweighted sum approach, the two constituting criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ can be fulfilled to an astonishing degree.

@@

Some simplifications in the computation of $\mathit{SSE_{kplusGeneralized}}$ can be arranged when ignoring the moment weights $\mathbf{w}$, e.g. by setting $w_t = 1$ ($t = 1, \dots, T$), leading to the *unweighted sum approach* for k-plus anticlustering. In this case, the additional dimension in $\mathbf{\hat{X}}_{T \times N \times P}$, representing distribution moments can be dropped; instead, the tensor is flattened into a two-dimensional matrix, where the variables representing the additional distribution moments are appended to the original data matrix $\mathbf{X}_{N \times P}$, resulting in the augmented matrix $\mathbf{X'}_{N \times \mathit{TP}}$:

@@

Thus, when setting the weights for all moments to 1, the k-plus criterion $\mathit{SSE_{kplusGeneralized}}$ has the same form as the standard k-means criterion $\mathit{SSE}$, but is based on different data. Here, computing the k-plus criterion is done by first creating the additional k-plus variables, appending them as new columns to the original data matrix, and then computing the $\mathit{SSE}$ on the augmented data. Using the unweighted sum approach, k-plus anticlustering therefore leaves the original k-means criterion unchanged.

@@

Two simulation studies were conducted to evaluate the new k-plus objective for anticlustering. A first small scale simulation compared different approaches for weight selection for the k-plus criterion $\mathit{SSE_{kplusGeneralized}}$. In a second large scale simulation study, the k-plus objective was compared to other popular anticlustering objectives, specifically, the diversity and the k-means objectives.

@@

In both simulation studies, the same local maximum search was applied to optimize the competing anticlustering objectives [Method LCW, @weitz1998]. Building on an initial random allocation, the algorithm proceeds by swapping elements between groups in such a way that each swap improves the objective criterion by the largest possible margin. That is, it starts with the first element and simulates all exchanges with elements that are currently assigned to a different group. After each exchange has been evaluated, the one exchange is realized that improves the objective function the most. No exchange is realized if no improvement is possible. The exchange process is repeated for each element. After that, the procedure restarts at the beginning and is repeated until an iteration through the entire data set no longer yields an improvement. To obtain better results, this local maximum search may be initialized multiple times [@spath1986]. In the simulation studies, five repetitions were used. Note that the local maximum search can be performed on the basis of any arbitrary initial grouping. In particular, the initial grouping directly determines the resulting group sizes, because the following exchange procedure never changes the sizes of the groups. Hence, while the data sets in the simulation studies were always partitioned into equal-sized groups, it is possible to specify different sized groups through the initial assignment, or if the data set cannot evenly be split, it is possible to ensure that the group sizes differ by 1 at most.

@@

In Simulation 1, the target was the optimization of $\mathit{SSE_{kplusGeneralized}}$ with $T = 4$, i.e. minimizing differences with regard to group means, variances, skewness and kurtosis. Six weighting schemes were compared. Two unweighted approaches were implemented by defining $\mathbf{w} = (1, 1, 1, 1)$. The unweighted standardization approach implemented a $z$-standardization of all columns---the original as well as the augmented k-plus variables---of the input data before the criterion $\mathit{SSE_{kplusGeneralized}}$ was optimized. In the second case, the unstandardised data matrix was used. Four weighted approaches were implemented by giving a tenfold increased weight to one of the four moments, respectively. That is, to compute $\mathit{SSE_{kplusGeneralized}}$ favouring similarity in means during the optimization process, I set $\mathbf{w} = (10, 1, 1, 1)$; the weights favouring similarity in variances, skewness and kurtosis, were defined as $\mathbf{w} = (1, 10, 1, 1)$, $\mathbf{w} = (1, 1, 10, 1)$, and $\mathbf{w} = (1, 1, 1, 10)$, respectively. The weighted approaches did not use a standardization of the input data.

@@

Table 2 displays the global simulation results aggregated across all conditions. Standard k-means performed best at minimizing discrepancy with regard to means ($\Delta \mathit{M}$), but did not outperform a random assignment with regard to all other distribution characteristics. Replicating @papenberg2020, k-means' performance with regard to similarity in standard deviations was even below that of a random assignment. K-plus anticlustering addressed $\Delta \mathit{M}$ almost as well as k-means, and at the same time minimized discrepancy with regard to the standard deviations ($\Delta \mathit{SD}$); both k-plus approaches strongly outperformed all other methods with regard to minimizing $\Delta \mathit{SD}$.

@@

apa_table(
  ARIs,
  caption = "Average pairwise ARI returned by the competing methods in Simulation 2.",
  note = "Each cell value is the result of averaging across 30,000 data points (10,000 data sets $\\times$ $K = 2$, $K = 3$ or $K = 4$).",
  escape = FALSE,
  align = c("l")
)

@@

The present paper focused on the new k-plus criterion for anticlustering, but only briefly considered the multicriterion nature of this criterion. A simple unweighted sum approach was used to optimize the k-plus objective, while more sophisticated algorithmic procedures exist to tackle multicriterion optimization problems [e.g., @brusco2009cross]. The sole focus on the objective function followed two reasonings. First, the applications and the simulations showed that the unweighted standardization approach yielded satisfying results that probably suffice in most applications. Moreover, it is possible that k-plus anticlustering has a fortunate solution space where many suitable partitions exist that fulfill multiple criteria to a satisfactory degree (as suggested by Figure 1). Still, more sophisticated approaches may yield theoretically improved results, as well as the opportunity to deliberately select among partitionings that may be preferred according to different k-plus objectives. @brusco2019 gave an example of how direct algorithms can be employed to optimize multiple criteria in anticlustering applications aiming for high within-group heterogeneity. Future research should investigate if k-plus anticlustering may profit from similar algorithmic treatment. Specifically, it should be investigated if the bicriterion approach by Brusco et al. can be transformed into a multicriterion algorithm for the k-plus objective. A second reason for focusing on the k-plus objective per se---independently from the algorithm that is used to optimize it---was to obtain a fair comparison to existing anticlustering objectives. In the simulation study, differences in performance could not have been interpreted if the objective function as well as the algorithm used to optimize it had differed between anticlustering methods. It was of particular interest to compare the k-plus objective with the popular diversity objective. @papenberg2020 compared k-means anticlustering and diversity anticlustering[^clusterediting] in a simulation study and concluded that the diversity criterion should be preferred because k-means does not equalize the spread of the distribution between groups. In contrast, maximizing the diversity leads to an appropriate balance between location and spread. Using the k-plus extension, however, the k-means criterion has regained attractivity: When opting for similarity in means and variances---and for increasing $N$, skewness and kurtosis as well---, k-plus anticlustering clearly outperformed diversity anticlustering. Future research should investigate whether this advantage also holds for other data distributions than normal. For example, it may be hypothesized that a generalized k-plus criterion is even more useful when applied to data that is not symmetric or contains outliers, because it is able to specifically incorporate the skewness and kurtosis as criteria in the anticlustering process.

As an additional outlook, it should be noted that because the covariance is defined as an expected value, the k-plus logic can be extended to minimize differences with regard to covariances, a possibility that was not investigated in the current paper. In this case, we have to compute an additional variable for each of pair of variables whose covariance should be similar between groups. In future research, the unweighted standardization approach presented here may serve as a starting point for investigating covariance similarity in anticlustering applications.

>>>>

@@

By maximizing the bicriterion objective $\mathit{SSE_{kplus}}$, we hope to obtain a partitioning where means as well as variances are similar between groups. The weights $w_1$ and $w_2$ can be chosen to adjust the relative importance of equalizing means and variances, respectively, during the optimization. Research on multiobjective optimization indicates that solutions oftentimes fail to satisfactorily address several criteria if the underlying objectives are in conflict rather than in alignment (Ferligoj & Batagelj, 1992). As $\mathit{SSE_{kplus}}$ is a novel criterion, it is so far unknown if the two constituting criteria are aligned. That is, it has yet to be demonstrated that maximizing $\mathit{SSE_{kplus}}$ can yield a partition where means as well as variances are similar between groups. A motivating example illustrates that aiming to simultaneously optimize similarity of means and variances may be feasible. I created 14 data points following a univariate normal distribution ($M = 15$, $SD = 3$) and then generated all `r n_partitions(14, 2)` possible ways to partition the 14 data points into $K = 2$ equal-sized groups. Figure 1 shows the scatterplot of the criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ across partitions. The partition that is represented by the squared symbol has the overall best highest $\mathit{SSE}$, but is outperformed by many other partitions with regard to $\mathit{SSE_{Var}}$. Similarly, the partition that maximizes $\mathit{SSE_{Var}}$ (represented by the triangular symbol) is not well-suited to optimize $\mathit{SSE}$ at the same time. Thus, when only considering one of the objectives, the mathematically optimal partitioning fails to establish groups that are similar both with regard to their means and variances. The diamond symbol represents the partition that maximizes the unweighted sum of $\mathit{SSE_{Var}}$ and $\mathit{SSE}$ (i.e., $w_1$ = $w_2$ = 1 in the computation of $\mathit{SSE_{kplus}}$), which yields `r df$kmeans[which.max(df$kplus)] / max(df$kmeans) * 100`% of the global maximum $\mathit{SSE}$, and `r df$kvar[which.max(df$kplus)] / max(df$kvar) * 100`% of the global maximum value of $\mathit{SSE_{Var}}$. Thus, even using a very simple unweighted sum approach, the two constituting criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ can be fulfilled to an astonishing degree.

@@

Some simplifications in the computation of $\mathit{SSE_{kplusGeneralized}}$ can be arranged when ignoring the moment weights $\mathbf{w}$ by setting $w_t = 1$ ($t = 1, \dots, T$); this simplification leads to the *unweighted sum approach* for k-plus anticlustering. In this case, the additional dimension in $\mathbf{\hat{X}}_{T \times N \times P}$, representing distribution moments can be dropped; instead, the tensor is flattened into a two-dimensional matrix, where the variables representing the additional distribution moments are appended to the original data matrix $\mathbf{X}_{N \times P}$, resulting in the augmented matrix $\mathbf{X'}_{N \times \mathit{TP}}$:

@@

Thus, when setting the weights for all moments to 1, the k-plus criterion $\mathit{SSE_{kplusGeneralized}}$ has the same form as the standard k-means criterion $\mathit{SSE}$. Here, computing the k-plus criterion is done by first creating the additional k-plus variables, appending them as new columns to the original data matrix, and then computing the $\mathit{SSE}$ on the augmented data. Using the unweighted sum approach, k-plus anticlustering therefore leaves the original k-means criterion unchanged.

@@

Two simulation studies were conducted to evaluate the new k-plus objective for anticlustering. A first small scale simulation compared different approaches for weight selection for the k-plus criterion $\mathit{SSE_{kplusGeneralized}}$. In a second large scale simulation study, the k-plus objective was compared to other anticlustering objectives striving for between-group similarity, specifically, the diversity and the k-means objectives.

@@

In both simulation studies, the same local maximum search was applied to optimize the competing anticlustering objectives [Method LCW, @weitz1998]. Building on an initial random allocation, the algorithm proceeds by swapping elements between groups in such a way that each swap improves the objective criterion by the largest possible margin. That is, it starts with the first element and simulates all exchanges with elements that are currently assigned to a different group. After each exchange has been evaluated, the one exchange is realized that improves the objective function the most. No exchange is realized if no improvement is possible. The exchange process is repeated for each element. After that, the procedure restarts at the beginning and is repeated until an iteration through the entire data set no longer yields an improvement. To obtain better results, this local maximum search may be initialized multiple times [@spath1986]. In the simulation studies, five repetitions were used. Note that the local maximum search can be performed on the basis of any arbitrary initial grouping. In particular, the initial grouping directly determines the size of the anticlusters, because the exchange procedure following the set-up never changes the sizes of the groups. While the data sets in the simulation studies were always partitioned into equal-sized groups, it is also possible to request different-sized groups or, if the data set cannot be split evenly, it is possible to ensure that the group sizes differ by 1 at most.

@@

In Simulation 1, the target was the optimization of $\mathit{SSE_{kplusGeneralized}}$ with $T = 4$, i.e. minimizing differences with regard to group means, variances, skewness and kurtosis. Six weighting schemes were compared. Two unweighted approaches were implemented by defining $\mathbf{w} = (1, 1, 1, 1)$. The unweighted standardization approach implemented a $z$-standardization of all columns---the original as well as the augmented k-plus variables---of the input data before the criterion $\mathit{SSE_{kplusGeneralized}}$ was optimized. In the second case, the unstandardised augmented data matrix was used. Four weighted approaches were implemented by giving a tenfold increased weight to one of the four moments, respectively. That is, to compute $\mathit{SSE_{kplusGeneralized}}$ favouring similarity in means during the optimization process, I set $\mathbf{w} = (10, 1, 1, 1)$; the weights favouring similarity in variances, skewness and kurtosis, were defined as $\mathbf{w} = (1, 10, 1, 1)$, $\mathbf{w} = (1, 1, 10, 1)$, and $\mathbf{w} = (1, 1, 1, 10)$, respectively. The weighted approaches did not use a standardization of the input data.

@@

Table 2 displays the global simulation results aggregated across all conditions. Standard k-means performed best at minimizing discrepancy with regard to means ($\Delta \mathit{M}$), but did not outperform a random assignment with regard to all other distribution characteristics. K-plus anticlustering addressed $\Delta \mathit{M}$ almost as well as k-means, and at the same time minimized discrepancy with regard to the standard deviations ($\Delta \mathit{SD}$); both k-plus approaches strongly outperformed all other methods with regard to minimizing $\Delta \mathit{SD}$.

@@

# remove diagonal
diag(ARIs) <- ""
ARIs <- ARIs[, -ncol(ARIs)]

apa_table(
  ARIs,
  caption = "Average pairwise ARI returned by the competing methods in Simulation 2.",
  note = "Each cell value is the result of averaging across 30,000 data points (10,000 data sets $\\times$ $K = 2$, $K = 3$ or $K = 4$).",
  escape = FALSE,
  align = c("r")
)

@@

The present paper focused on the new k-plus criterion for anticlustering, but only briefly considered the multicriterion nature of this criterion. A simple unweighted sum approach was used to optimize the k-plus objective, while more sophisticated algorithmic procedures exist to tackle multicriterion optimization problems [e.g., @brusco2009cross]. The sole focus on the objective function followed two reasonings. First, the applications and the simulations showed that the unweighted standardization approach yielded satisfying results that probably suffice in most applications. Moreover, it is possible that k-plus anticlustering has a fortunate solution space where many suitable partitions exist that fulfill multiple criteria to a satisfactory degree (as suggested by Figure 1). Still, more sophisticated approaches may yield theoretically improved results, as well as the opportunity to deliberately select among partitionings that may be preferred according to different k-plus objectives. @brusco2019 gave an example of how direct algorithms can be employed to optimize multiple criteria in anticlustering applications aiming for high within-group heterogeneity. Future research should investigate if k-plus anticlustering may profit from similar algorithmic treatment. Specifically, it is of high interest to investigate if the bicriterion approach by Brusco et al. can be transformed into a multicriterion algorithm for the k-plus objective. A second reason for focusing on the k-plus objective per se---independently from the algorithm that is used to optimize it---was to obtain a fair comparison to existing anticlustering objectives. In the simulation study, differences in performance could not have been interpreted if the objective function as well as the algorithm used to optimize it had differed between anticlustering methods. It was of particular interest to compare the k-plus objective with the popular diversity objective. @papenberg2020 compared k-means anticlustering and diversity anticlustering[^clusterediting] in a simulation study and concluded that the diversity criterion should be preferred because k-means does not equalize the spread of the distribution between groups. In contrast, maximizing the diversity leads to an appropriate balance between location and spread. Using the k-plus extension, however, the k-means criterion has regained attractivity: When opting for similarity in means and variances---and for increasing $N$, skewness and kurtosis as well---, k-plus anticlustering clearly outperformed diversity anticlustering. Future research should investigate whether this advantage also holds for other data distributions than normal. For example, it may be hypothesized that the generalized k-plus criterion is even more useful when applied to data that is not symmetric or contains outliers, because it is able to explicitly incorporate the skewness and kurtosis as criteria in the anticlustering process.

As an additional outlook, it should be noted that because the covariance is defined as an expected value, the k-plus logic can be extended to minimize differences with regard to covariances, a possibility that was not investigated in the current paper. To take into account covariances, we have to compute an additional variable for each of pair of variables whose covariance should be similar between groups. The unweighted standardization approach presented here may serve as a starting point for investigating covariance similarity in anticlustering applications in the future.
